---
title: "EDA"
author: "Shina Min"
output: github_document
---

```{r set up}
library(tidyverse)
library(ggridges)
```


## Load the weather_df 
```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = recode(
      id, 
      USW00094728 = "CentralPark_NY", 
      USW00022534 = "Molokai_HI",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) |>
  select(name, id, everything())
## using cached file: /Users/jeffgoldsmith/Library/Caches/org.R-project.R/R/rnoaa/noaa_ghcnd/USW00094728.dly
## date created (size, mb): 2023-09-19 15:41:55.07359 (8.524)
## file min/max dates: 1869-01-01 / 2023-09-30
## using cached file: /Users/jeffgoldsmith/Library/Caches/org.R-project.R/R/rnoaa/noaa_ghcnd/USW00022534.dly
## date created (size, mb): 2023-09-25 10:06:23.827176 (3.83)
## file min/max dates: 1949-10-01 / 2023-09-30
## using cached file: /Users/jeffgoldsmith/Library/Caches/org.R-project.R/R/rnoaa/noaa_ghcnd/USS0023B17S.dly
## date created (size, mb): 2023-09-19 15:42:03.139582 (0.994)
## file min/max dates: 1999-09-01 / 2023-09-30

weather_df
## # A tibble: 2,190 × 6
##    name           id          date        prcp  tmax  tmin
##    <chr>          <chr>       <date>     <dbl> <dbl> <dbl>
##  1 CentralPark_NY USW00094728 2021-01-01   157   4.4   0.6
##  2 CentralPark_NY USW00094728 2021-01-02    13  10.6   2.2
##  3 CentralPark_NY USW00094728 2021-01-03    56   3.3   1.1
##  4 CentralPark_NY USW00094728 2021-01-04     5   6.1   1.7
##  5 CentralPark_NY USW00094728 2021-01-05     0   5.6   2.2
##  6 CentralPark_NY USW00094728 2021-01-06     0   5     1.1
##  7 CentralPark_NY USW00094728 2021-01-07     0   5    -1  
##  8 CentralPark_NY USW00094728 2021-01-08     0   2.8  -2.7
##  9 CentralPark_NY USW00094728 2021-01-09     0   2.8  -4.3
## 10 CentralPark_NY USW00094728 2021-01-10     0   5    -1.6
## # ℹ 2,180 more rows

```

```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2023-12-31") |>
  mutate(
    name = recode(
      id, 
      USW00094728 = "CentralPark_NY", 
      USW00022534 = "Molokai_HI",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10,
    month = lubridate::floor_date(date, unit = "month")) %>%
  select(name, id, everything())

weather_df
```


Introducing 

## `group_by`

```{r}
weather_df %>%
  group_by(name, month)
```


## counting things 

count month / name observations

```{r}
weather_df %>%
  group_by(name, month) %>%
  summarize(n_obs = n())
```

we can use `count`

```{r}
weather_df %>%
  count(name, month, name = "n_obs")
```

**NEVER** use base R's `table` function

```{r, eval=FALSE}
weather_df %>%
  pull(month) %>%
  table()
```


other helpful counters

```{r}
weather_df %>%
  group_by(month) %>%
  summarize(
    nobs = n(),
    n_days = n_distinct(date))
```


## A digression on 2x2 tables

```{r}
weather_df %>%
  filter(name !="Molokai_HI") %>%
  mutate(
    cold = case_when(
      tmax < 5 ~ "cold",
      tmax >= 5 ~ "not cold",
      TRUE      ~ ""     
      )
  ) %>% 
  group_by(name, cold) %>% 
  summarize(count = n())
```

```{r}
weather_df %>%
  filter(name !="Molokai_HI") %>%
  mutate(
    cold = case_when(
      tmax < 5 ~ "cold",
      tmax >= 5 ~ "not cold",
      TRUE      ~ ""     
      )
  ) %>%
  janitor::tabyl(name, cold)
```


## General summaries 

You can do lots of summaries.

```{r}
weather_df %>%
  group_by(month) %>%
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE),
    mean_prcp = mean(prcp, na.rm = TRUE),
    median_tmin = median(tmin, na.rm = TRUE)
  )
```

This is a dataframe! So we can make a plot!

```{r}
weather_df %>%
  group_by(name, month) %>%
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE),
    mean_prcp = mean(prcp, na.rm = TRUE),
    median_tmin = median(tmin, na.rm = TRUE)
  ) %>%
  ggplot(aes(x = month, y = mean_tmax, color = name)) +
  geom_point() +
  geom_line()
```

Suppose you want to summarize many columns.

```{r}
weather_df %>%
  group_by(name, month) %>%
  summarize(across(prcp:tmin, mean))
```

Reminder: sometimes your results are easier to read in another format

```{r}
weather_df %>%
  group_by(name, month) %>%
  summarize(mean_tmax = mean(tmax)) %>%
  pivot_wider(
    names_from = name,
    values_from = mean_tmax
  ) %>%
  knitr::kable(digits = 1)
```

## `group_by` and `mutate`

```{r}
weather_df %>%
  group_by(name) %>% 
  mutate(
    mean_tmax = mean(tmax, na.rm = TRUE),
    centered_tmax = tmax - mean_tmax
  ) %>%
  ggplot(aes(x = date, y = centered_tmax, color = name)) +
  geom_point()
```


what about window functions?

Ranking...
```{r}
weather_df %>%
  group_by(name, month) %>%
  mutate(temp_rank = min_rank(tmax))
```

Lag

```{r}
weather_df %>%
  group_by(name) %>%
  mutate(temp_change = tmax - lag(tmax)) %>%
  summarize(
    temp_change_max = max(temp_change, na.rm = TRUE),
    temp_change_sd = sd(temp_change, na.rm = TRUE)
  )
```

