---
title: "Visualization"
subjtitle: "Part I"
output: github_document
---

```{r set up}
library(tidyverse)
library(ggridges)
```


## Load the weather_df 
```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = recode(
      id, 
      USW00094728 = "CentralPark_NY", 
      USW00022534 = "Molokai_HI",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) |>
  select(name, id, everything())
## using cached file: /Users/jeffgoldsmith/Library/Caches/org.R-project.R/R/rnoaa/noaa_ghcnd/USW00094728.dly
## date created (size, mb): 2023-09-19 15:41:55.07359 (8.524)
## file min/max dates: 1869-01-01 / 2023-09-30
## using cached file: /Users/jeffgoldsmith/Library/Caches/org.R-project.R/R/rnoaa/noaa_ghcnd/USW00022534.dly
## date created (size, mb): 2023-09-25 10:06:23.827176 (3.83)
## file min/max dates: 1949-10-01 / 2023-09-30
## using cached file: /Users/jeffgoldsmith/Library/Caches/org.R-project.R/R/rnoaa/noaa_ghcnd/USS0023B17S.dly
## date created (size, mb): 2023-09-19 15:42:03.139582 (0.994)
## file min/max dates: 1999-09-01 / 2023-09-30

weather_df
## # A tibble: 2,190 × 6
##    name           id          date        prcp  tmax  tmin
##    <chr>          <chr>       <date>     <dbl> <dbl> <dbl>
##  1 CentralPark_NY USW00094728 2021-01-01   157   4.4   0.6
##  2 CentralPark_NY USW00094728 2021-01-02    13  10.6   2.2
##  3 CentralPark_NY USW00094728 2021-01-03    56   3.3   1.1
##  4 CentralPark_NY USW00094728 2021-01-04     5   6.1   1.7
##  5 CentralPark_NY USW00094728 2021-01-05     0   5.6   2.2
##  6 CentralPark_NY USW00094728 2021-01-06     0   5     1.1
##  7 CentralPark_NY USW00094728 2021-01-07     0   5    -1  
##  8 CentralPark_NY USW00094728 2021-01-08     0   2.8  -2.7
##  9 CentralPark_NY USW00094728 2021-01-09     0   2.8  -4.3
## 10 CentralPark_NY USW00094728 2021-01-10     0   5    -1.6
## # ℹ 2,180 more rows

```


## Scatterplot!!

Create my first scatterplot ever.

```{r}
ggplot(weather_df, aes(x = tmin, y = tmax)) +
  geom_point()
```

New apporach, same plot.

```{r}
weather_df %>%
  ggplot(aes(x = tmin, y = tmax)) +
  geom_point()
```

Save and edit a plot object.

```{r}
weather_plot =
  weather_df %>%
  ggplot(aes(x = tmin, y = tmax))
  
weather_plot + geom_point()

```

## ADvanced scatterplot ...

Start with the same one and make it fancy!

```{r}
weather_df %>%
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point() +
  geom_smooth(se = FALSE)

```

What about the `aes` placement ..?

```{r}
weather_df %>%
  ggplot(aes(x = tmin, y = tmax)) +
  geom_point(aes(color = name)) +
  geom_smooth()
```

Let's facet some things!!

```{r}
weather_df %>%
  ggplot(aes(x = tmin, y = tmax, alpha = tmin, color = name)) +
  geom_point() +
  geom_smooth(se = FALSE, size = 2) +
  facet_grid(. ~ name)
```

Let's combine some elements and try a new plot.

```{r}
weather_df %>%
  ggplot(aes(x = date, y = tmax, color = name)) +
  geom_point(aes(size = prcp), alpha = 0.5) +
  geom_smooth(se = FALSE) +
  facet_grid(. ~ name)

```



## Some small notes

How many geoms have to exist? 

```{r}
weather_df %>%
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = 0.5) +
  geom_smooth(se = FALSE)
```

You can use a `neat` geom!

```{r}
weather_df %>%
  ggplot(aes(x = tmin, y = tmax)) +
  geom_density2d() +
  geom_point(alpha = 0.3)
```

## Univariate plots

Histograms are really great.

```{r}
weather_df %>%
  ggplot(aes(x = tmin)) +
  geom_histogram()
```

Can we add color...?

```{r}
weather_df %>%
  ggplot(aes(x = tmin)) +
  geom_histogram(position = "dodge") +
  facet_grid(. ~ name)
```

Let's try a new geometry!

```{r}
weather_df %>%
  ggplot(aes(x = tmin, fill = name)) +
  geom_density(alpha = 0.3)
```


What about box plots??

```{r}
weather_df %>%
  ggplot(aes(x = name, y = tmin)) +
  geom_boxplot()
```

Trendy plots :-)

```{r}
weather_df %>%
  ggplot(aes(x = name, y = tmin, fill = name)) +
  geom_violin(alpha = 0.5) +
  stat_summary()
```

Ridge plots -- the most popular plot of 2017

```{r}
weather_df %>%
  ggplot(aes(x = tmin, y = name)) +
  geom_density_ridges()
```

